Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2018-12-04 Manuscript/manuscript.Rmd
manuscript edits

Using each genomic element, we created three types of predictors that were used as features to analyze their relationships with TAD boundaries in downstream modelling (Figure 2). The predictors consisted of two different types of overlaps, and a distance type. They are described in detail below:

   1.	Overlap Counts (OC): For each genomic bin, the total number of overlapped regions between a functional genomic element and a bin was calculated. This was defined as the overlap counts (OC) and is illustrated in the left part of Figure 2. Within each bin (marked by the blue dashed lines), the total number of overlapped regions of the genomic element (represented by the horizontal green segments) are tallied and marked below. In the case of no overlaps the value is given as 0.

   2.	Overlap Percent (OP): For each genomic bin, the percent overlap between regions defined by a particular functional element and a bin was calculated. From the middle diagram of Figure 2, the feature width was defined as the number of base pairs that overlapped with the genomic bin. The percent was then calculated by dividing the feature width with the bin width. If multiple overlaps existed, the feature width was defined as the sum of the total number of base pairs that overlapped with a particular bin. Note that the bin width remained constant for a given resolution (either 5kb, 25kb, 50kb, or 100kb). Similar to the OC predictor type, bins with no overlaps were given to be 0.

   3.	Distance: For each genomic bin, the distance (in base pairs) from the center of the bin to the center of the nearest region of a respective functional element was calculated. The right most diagram of Figure 2 depicts an example of how distance was calculated. The vertical green segments represent the centers of the respective genomic element. Each distance, $D_{i}$, represented by the blue brackets are the number of base pairs from the center of the bins to the center of the nearest genomic elemental region.

@@

Figure 3 details the ensemble framework proposed and evaluated in this study. For each resolution and predictor type chosen, the data was split into training to testing sets. A variable reduction technique was then performed. The feature space was reduced and a resampling technique was applied. Lastly, a classification algorithm in the form of a random forest was performed and validated. Details of each step in the framework are provided below.

@@

Given a particular resolution, the full data set was composed of genomic bins that either overlapped with a flanked TAD boundary (minority set) or did not (majority set). Figure 3 highlights the severe class imbalance present in both the full data set. The data was then split into a 7:3 ratio of training to testing sets. Each training and testing set was composed of a similar majority to minority class ratio as the full data set, which is represented by the segmented regions in Figure 3. The training set was then used to perform all downstream modeling. The same test set was held separate and used to validate all models.

@@

We evaluated the use of two variable reduction algorithms, the LASSO and Elastic-Net regularizations. Each algorithm was performed prior to re-sampling. For the LASSO, the $\lambda$ penalization term was tuned over a grid of 10 values on an exponential scale ranging from 0.01 to 10. The additional $\alpha$ term for the Elastic-Net was tuned over 10 equa-distant values from 0 to 1.

@@

After choosing one of the aforementioned data resolutions, predictor types, and variable reduction algorithms, we then evaluated three re-sampling techniques used to create balanced classes.  For each resampling technique, the majority class is defined as the set of genomic bins that did not include a TAD boundary, while the minority class is defined as the set of bins that included a TAD boundary. The following details the re-sampling techniques considered in this paper:

@@

A classification algorithm in the form of a random forest was performed and validated using the test set. 10-fold cross validation was used to reduced bias due to random dataset generation. The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. Once the model was implemented, it was assessed using the test set, and performance metrics were then recorded.

>>>>

Using each genomic element, we created three types of predictors that were used as features to analyze their relationships with TAD boundaries in downstream modelling. The predictors consisted of two different types of overlaps, and a distance type. They are described in detail below:

   1.	Overlap Counts (OC): For each genomic bin, the total number of overlapped regions between a functional genomic element and a bin was calculated. This was defined as the overlap counts (OC) and is illustrated in the left part of Figure 2. Within each bin (marked by the blue dashed lines), the total number of overlapped regions of the genomic element (represented by the horizontal green segments) are tallied and marked below. In the case of no overlaps the value is given as 0 (Figure 2 A).

   2.	Overlap Percent (OP): For each genomic bin, the percent overlap between regions defined by a particular functional element and a bin was calculated. From the middle diagram of Figure 2, the feature width was defined as the number of base pairs that overlapped with the genomic bin. The percent was then calculated by dividing the feature width with the bin width. If multiple overlaps existed, the feature width was defined as the sum of the total number of base pairs that overlapped with a particular bin. Note that the bin width remained constant for a given resolution (either 5kb, 25kb, 50kb, or 100kb). Similar to the OC predictor type, bins with no overlaps were given to be 0 (Figure 2 B).

   3.	Distance: For each genomic bin, the distance (in base pairs) from the center of the bin to the center of the nearest region of a respective functional element was calculated. The right most diagram of Figure 2 depicts an example of how distance was calculated. The vertical green segments represent the centers of the respective genomic element. Each distance, $D_{i}$, represented by the blue brackets are the number of base pairs from the center of the bins to the center of the nearest genomic elemental region (Figure 2 C).

@@

An ensemble of models was established for the purposes of directly comparing the performances of different parameters in the framework. Each system in the framework is assumed to be composed of a model constructed from a particular data resolution, with a particular predictor type as the feature space. Details of each step in the framework are provided below.

@@

The data set used for modeling was composed of the full set of genomic bins that either overlapped with a flanked TAD boundary or did not. These two classes made up the majority and minority sets (Figure 3 A). The data was then split into a 7:3 ratio of training to testing sets (Figure 3 B). Each training and testing set was composed of a similar majority to minority class ratio as the full data set (Figure 3 C). The training set was then used to perform all downstream modeling.

@@

We evaluated the use of two variable reduction algorithms, the LASSO and Elastic-Net regularizations (Figure 3 D). Each algorithm was performed prior to re-sampling. For the LASSO, the $\lambda$ penalization term was tuned over a grid of 10 values on an exponential scale ranging from 0.01 to 10. The additional $\alpha$ term for the Elastic-Net was tuned over 10 equa-distant values from 0 to 1.

@@

We then evaluated three re-sampling techniques (in addition to no re-sampling) used to create balanced classes (Figure 3 E). They are detailed below:

@@

A classification algorithm in the form of a random forest was performed and validated using the same test set for all models in the ensemble (Figure 3 F). 10-fold cross validation was used to reduced bias due to random dataset generation. The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. Once the model was implemented, it was assessed using the test set, and performance metrics were then recorded.
